State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-18 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edits to abstract in intro of paper, added kernel figure

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used as the parametric form of the model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using modelled sensors and the field model. We introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model to facilitate state and parameter estimation. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. We use the reduced model for the estimation that consists of a two-stage algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results illustrate the cost of the model reduction with a loss of the high spatial frequency information. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

\section{Introduction} This paper presents a theoretical framework for generating macroscopic neural field mathematical models from patient-specific data. There has been much interest in generating physiologically plausible neural field models to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is this scale at which pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested. Mathematical neural field models provide insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see \cite{Deco2008,David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999,RENNIE2000}, epileptic seizure generation \cite{DaSilva2003,Suffczynski2004,Wendling2005} and insights into other pathologies \cite{Moran2008,Schiff2009} that would be difficult to gain from experimental data alone.

@@

This paper extends the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for higher order systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been reported (to the best of the authors knowledge). Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development here was to represent the spatiotemporal system as a standard state-space model with an order independent of the number of observation locations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of the technique to neural fields.

@@

\dean{show figure with the Mexican hat here.}

>>>>

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite dimensional state-space model. Spatial frequency analysis methods are introduced for model selection, by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. We use the reduced model to estimate the state and parameters of the non-reduced model. The estimation procedure consists of a two-stage algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation, and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic time constant with the proposed framework. The results also illustrate the cost of the model reduction with a loss of the high spatial frequency information. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

\section{Introduction} This paper presents a theoretical framework for generating patient-specific macroscopic neural field mathematical models from data. There has been much interest in generating physiologically plausible neural field models to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is this scale at which pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested. Mathematical neural field models provide insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see \cite{Deco2008,David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999,RENNIE2000}, epileptic seizure generation \cite{DaSilva2003,Suffczynski2004,Wendling2005} and insights into other pathologies \cite{Moran2008,Schiff2009} that would be difficult to gain from experimental data alone.

@@

This paper extends the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for higher order systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been reported (to the best of the authors knowledge). Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development from this recent paper was to represent the spatiotemporal system as a standard state-space model, with the number of states independent of the number of observation locations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of the technique to neural fields.

@@

\begin{figure}[th]\label{fig:experimental design}
   	\begin{center}
   		\includegraphics[width=0.3\textwidth]{./Graph/2D_kernel.pdf} 
   	\end{center}
   	\caption{Kernel centred at zero as a function of space.} 
   \end{figure}
